---
title: "Before you start: setup and your first PLINK command"
---

This page gets your computer ready and walks you through your **very first command**. By the
end you will have downloaded the demo dataset, installed PLINK, and run a quick test that
confirms everything works — computing allele frequencies with `--freq`.

::: {.callout-note title="What you will need"}
- A computer running **Windows, macOS or Linux**
- The **demo dataset** (download in Step 1)
- **PLINK 2** (download in Step 2)
- *Recommended:* **R** and **RStudio**, to run the analysis and plotting steps later in the guide

You do **not** need any programming experience. Every command is given in full.
:::

## A note on the command line {#sec-terminal}

Most steps in this guide are typed into a **terminal** (also called a "command line" or
"console") — a window where you type a command, press <kbd>Enter</kbd>, and the program runs.

- **Windows:** open the *Command Prompt* (search for `cmd`) or *PowerShell*.
- **macOS:** open the *Terminal* app (in *Applications → Utilities*).
- **Linux:** open your *Terminal*.
- **Inside RStudio:** there is a *Terminal* tab next to the *Console* tab — you can use that too.

Throughout the guide, lines you type into the terminal look like this:

```bash
plink2 --version
```

You copy the command, paste it into your terminal, and press <kbd>Enter</kbd>. That's it.

---

## Step 1 — Download the demo dataset {#sec-download}

The demo dataset is called **`pdac_demo`**. Like all PLINK datasets, it is made of
**three files that always travel together** and must stay in the same folder:

| File | What it contains |
|------|------------------|
| `pdac_demo.bed` | the genotypes (the actual genetic data, in compact binary form) |
| `pdac_demo.bim` | the list of variants (one row per SNP: chromosome, position, alleles) |
| `pdac_demo.fam` | the list of individuals (one row per person: ID, sex, case/control status) |

::: {.callout-important}
Never rename or move just one of the three files. PLINK always looks for all three together
using the shared name (`pdac_demo`).
:::

You can download them in either of two ways.

::: {.panel-tabset}

### Point and click (easiest)

Open the release page and download the three files, or click each link directly:

- [`pdac_demo.bed`](https://github.com/mgentiluomo/how-to-gwas-pdac/releases/download/v0.1-data/pdac_demo.bed) — the genotypes (≈150 MB)
- [`pdac_demo.bim`](https://github.com/mgentiluomo/how-to-gwas-pdac/releases/download/v0.1-data/pdac_demo.bim) — the list of variants
- [`pdac_demo.fam`](https://github.com/mgentiluomo/how-to-gwas-pdac/releases/download/v0.1-data/pdac_demo.fam) — the list of individuals

Then move all three files into your working folder.
(Full release page: <https://github.com/mgentiluomo/how-to-gwas-pdac/releases/tag/v0.1-data>)

> **Windows tip:** make sure the browser did not add a hidden `.txt` at the end
> (e.g. `pdac_demo.fam.txt`). The names must be exactly `pdac_demo.bed`, `.bim`, `.fam`.

### In R / RStudio

Run this in the RStudio *Console* — it downloads the three files straight into your folder,
the same way on Windows, macOS and Linux:

```{r}
#| eval: false
base <- "https://github.com/mgentiluomo/how-to-gwas-pdac/releases/download/v0.1-data/"
for (f in c("pdac_demo.bed", "pdac_demo.bim", "pdac_demo.fam")) {
  download.file(paste0(base, f), destfile = f, mode = "wb")
}
list.files(pattern = "pdac_demo")
```

### Command line

```bash
# macOS / Linux
for f in pdac_demo.bed pdac_demo.bim pdac_demo.fam; do
  curl -L -O "https://github.com/mgentiluomo/how-to-gwas-pdac/releases/download/v0.1-data/$f"
done
```

```powershell
# Windows PowerShell
$base = "https://github.com/mgentiluomo/how-to-gwas-pdac/releases/download/v0.1-data/"
foreach ($f in "pdac_demo.bed","pdac_demo.bim","pdac_demo.fam") {
  Invoke-WebRequest -Uri "$base$f" -OutFile $f
}
```

:::

::: {.callout-tip title="Pick one working folder and stay in it"}
Create a single folder for this tutorial — for example `gwas_tutorial` — and put the three data
files there. Do all your work inside it. This avoids almost every "file not found" problem
beginners run into.
:::

### Check the files arrived

In your terminal, move into the folder and list its contents:

```bash
cd path/to/gwas_tutorial      # replace with the real path to your folder
ls pdac_demo.*      # on Windows Command Prompt use: dir pdac_demo.*
```

You should see exactly three files: `.bed`, `.bim` and `.fam`.

---

## Step 2 — Install PLINK 2 {#sec-plink}

**PLINK** is the workhorse program for genotype data. We use **PLINK 2** (the command is
`plink2`). It is a single program file — there is **no installer**: you download it, unzip it,
and you are done.

1. Go to the official download page: <https://www.cog-genomics.org/plink/2.0/>
2. Download the build for your operating system (Linux / macOS / Windows).
3. **Unzip** it. Inside you will find the program file: `plink2` (macOS/Linux) or
   `plink2.exe` (Windows).

::: {.callout-tip title="The simplest setup"}
Put the `plink2` (or `plink2.exe`) file **in the same folder as your data**
(`gwas_tutorial`). Then you can run it as `./plink2` (macOS/Linux) or `plink2` (Windows) without
worrying about installation paths. More advanced users can instead add PLINK to their system
`PATH`.
:::

::: {.panel-tabset}

#### macOS / Linux

If you placed `plink2` in your working folder, you call it with `./plink2`. The first time on
macOS you may need to allow it to run:

```bash
chmod +x plink2          # make the file executable
./plink2 --version
```

#### Windows

Open *Command Prompt*, move to your folder, and call it as `plink2`:

```bash
cd path\to\gwas_tutorial
.\plink2 --version
```

:::

::: {.callout-note title="`plink2` vs `./plink2`"}
For the rest of this guide we write the command simply as `plink2`. On macOS/Linux, if PLINK is
in your current folder rather than installed system-wide, type `./plink2` instead. Everything
after the program name is identical.
:::

---

## Step 3 — Check that PLINK runs {#sec-version}

Before doing anything with the data, confirm PLINK starts at all:

```bash
.\plink2 --version
```

You should see a single line with the version and build date, similar to:

```
PLINK v2.0.0-a.6.9 64-bit (1 Mar 2025)
```

If you see a version line, PLINK works. If instead you get an error like
`command not found` or `'plink2' is not recognized`, jump to
[Troubleshooting](#sec-troubleshooting) below.

---

## Step 4 — Your first command: allele frequencies with `--freq` {#sec-freq}

Now the real test. We ask PLINK to read the demo dataset and compute the **allele frequency** of
every variant — that is, how common each allele is across the sample. This is the simplest
possible operation that touches the actual genotypes, so it is a perfect "does everything work?"
check.

```bash
.\plink2 --bfile pdac_demo --freq --out pdac_demo_freq
```

Let's read that command piece by piece:

| Part | Meaning |
|------|---------|
| `plink2` | run the PLINK 2 program |
| `--bfile pdac_demo` | load the dataset whose three files share this name (`.bed`/`.bim`/`.fam`) |
| `--freq` | compute the allele frequencies |
| `--out pdac_demo_freq` | name the output files with this prefix |

<span class="codetag">CODE 0-01: first PLINK run — allele frequencies — see /getting_started/</span>

When it finishes, PLINK prints a short log to the screen and creates a new file:
**`pdac_demo_freq.afreq`**. The on-screen log ends with something like:

```
... 
430000 variants loaded from pdac_demo.bim.
1461 samples (712 females, 749 males; ...) loaded from pdac_demo.fam.
--freq: Allele frequencies written to pdac_demo_freq.afreq .
```

If you see a line saying the allele frequencies were **written**, the test has passed: PLINK can
read your data and compute on it. 🎉

---

## Step 5 — Did it work? Look at the result {#sec-check}

Open the first few lines of the output file to see what PLINK produced:

```bash
head pdac_demo_freq.afreq    # on Windows use: more pdac_demo_freq.afreq
```

You should see a table with one row per variant, like this:

```
#CHROM  ID                REF  ALT  ALT_FREQS   OBS_CT
1       1:752566:G:A      G    A    0.831621    2922
1       1:776546:A:C      A    C    0.142847    2922
1       1:832918:T:C      T    C    0.067419    2922
1       1:842013:T:G      T    G    0.219370    2922
...
```

What the columns mean:

| Column | Meaning |
|--------|---------|
| `#CHROM` | chromosome |
| `ID` | the variant's identifier |
| `REF` / `ALT` | the reference and alternate alleles |
| `ALT_FREQS` | frequency of the **ALT** allele (a number between 0 and 1) |
| `OBS_CT` | number of observed alleles (≈ 2 × number of genotyped individuals) |

That `ALT_FREQS` column is the answer to our test: **PLINK computed a frequency for every
variant.**

### Optional: read it in R

If you have R/RStudio set up, you can load the same file and take a quick look. This also
introduces the pattern we use throughout the guide — PLINK produces a table, R reads and
summarises it:

```{r}
#| eval: false
# read the allele-frequency table PLINK just wrote
freq <- read.table("pdac_demo_freq.afreq",
                   header = TRUE, comment.char = "", sep = "\t",
                   check.names = FALSE)

# how many variants, and the first rows
nrow(freq)
head(freq)

# minor allele frequency (MAF) = the smaller of ALT_FREQS and 1 - ALT_FREQS
freq$MAF <- pmin(freq$ALT_FREQS, 1 - freq$ALT_FREQS)

# a quick picture of the frequency spectrum
hist(freq$MAF, breaks = 50,
     main = "Minor allele frequency distribution",
     xlab = "MAF")
```

A typical genotyping array shows many common variants and a tail of rarer ones, so the histogram
leans toward the lower MAF values. If R prints a number of variants and draws a histogram, your
R setup works too — and you are ready for the analysis sections.

::: {.callout-note title="Why we don't re-run PLINK every time the website is built"}
Real GWAS commands (especially on full datasets) can be slow and run on a dedicated server.
In this guide, PLINK commands are shown for **you** to run on your own machine; the website does
not re-execute them. Lightweight R steps (a plot, a small table) are run once and cached, so the
site stays fast and reproducible.
:::

---

## Troubleshooting {#sec-troubleshooting}

::: {.callout-warning collapse="true" title="`command not found` / `'plink2' is not recognized`"}
The terminal cannot find the PLINK program. Either you are not in the folder where `plink2`
lives, or it is not on your `PATH`.

- Make sure you have `cd`-ed into your working folder.
- On macOS/Linux, call it as `./plink2` (note the `./`).
- Check the file is really there: `ls plink2` (macOS/Linux) or `dir plink2.exe` (Windows).
:::

::: {.callout-warning collapse="true" title="`Error: Failed to open pdac_demo.bed`"}
PLINK cannot find the data files.

- Confirm all **three** files (`.bed`, `.bim`, `.fam`) are in the current folder and share the
  exact name `pdac_demo`.
- Confirm you typed the name **without** an extension after `--bfile` (PLINK adds `.bed`/`.bim`/`.fam`
  itself).
- Check you are in the right folder with `pwd` (macOS/Linux) or `cd` with no arguments (Windows).
:::

::: {.callout-warning collapse="true" title="macOS: \"plink2 cannot be opened because the developer cannot be verified\""}
This is macOS Gatekeeper. Either right-click the `plink2` file → *Open* once to allow it, or run
`xattr -d com.apple.quarantine plink2` in the terminal, then try again.
:::

---

## What's next

You now have a working setup and have confirmed PLINK can read the demo data. The next sections
of the guide build directly on this:

- **Quality control (Section 1B)** — cleaning the raw genotypes before any analysis.
- **Population stratification (Section 2)** — detecting and correcting for ancestry.

*(These sections will be linked here as they are added.)*
