---
title: "Before you start: setup and your first PLINK command"
---

This page gets your computer ready and walks you through your **very first command**. By the
end you will have downloaded the demo dataset, installed PLINK, and run a quick test that
confirms everything works — computing allele frequencies with `--freq`.

::: {.callout-note title="What you will need"}
- A computer running **Windows, macOS or Linux**
- The **demo dataset** (download in Step 1)
- **PLINK 2** (download in Step 2)
- *Recommended:* **R** and **RStudio**, to run the analysis and plotting steps later in the guide

You do **not** need any programming experience. Every command is given in full.
:::

## A note on the command line {#sec-terminal}

Most steps in this guide are typed into a **terminal** (also called a "command line" or
"console") — a window where you type a command, press <kbd>Enter</kbd>, and the program runs.

- **Windows:** open *PowerShell* or the *Command Prompt* (search for `cmd`).
- **macOS:** open the *Terminal* app (in *Applications → Utilities*).
- **Linux:** open your *Terminal*.
- **Inside RStudio:** there is a *Terminal* tab next to the *Console* tab — you can use that too.

Throughout the guide, lines you type into the terminal look like this:

```bash
plink2 --version
```

You copy the command, paste it into your terminal, and press <kbd>Enter</kbd>. That's it.

---

## Step 1 — Download the demo dataset {#sec-download}

The demo dataset is called **`pdac_demo`**. Like all PLINK datasets, it is made of
**three files that always travel together** and must stay in the same folder:

| File | What it contains |
|------|------------------|
| `pdac_demo.bed` | the genotypes (the actual genetic data, in compact binary form) |
| `pdac_demo.bim` | the list of variants (one row per SNP: chromosome, position, alleles) |
| `pdac_demo.fam` | the list of individuals (one row per person: ID, sex, case/control status) |

::: {.callout-important}
Never rename or move just one of the three files. PLINK always looks for all three together
using the shared name (`pdac_demo`).
:::

You can download them in either of two ways.

::: {.panel-tabset}

### Point and click (easiest)

Open the release page and download the three files, or click each link directly:

- [`pdac_demo.bed`](https://github.com/mgentiluomo/how-to-gwas-pdac/releases/download/v0.1-data/pdac_demo.bed) — the genotypes (≈150 MB)
- [`pdac_demo.bim`](https://github.com/mgentiluomo/how-to-gwas-pdac/releases/download/v0.1-data/pdac_demo.bim) — the list of variants
- [`pdac_demo.fam`](https://github.com/mgentiluomo/how-to-gwas-pdac/releases/download/v0.1-data/pdac_demo.fam) — the list of individuals

Then move all three files into your working folder.
(Full release page: <https://github.com/mgentiluomo/how-to-gwas-pdac/releases/tag/v0.1-data>)

> **Windows tip:** make sure the browser did not add a hidden `.txt` at the end
> (e.g. `pdac_demo.fam.txt`). The names must be exactly `pdac_demo.bed`, `.bim`, `.fam`.

### In R / RStudio

Run this in the RStudio *Console* — it downloads the three files straight into your folder,
the same way on Windows, macOS and Linux:

```{r}
#| eval: false
base <- "https://github.com/mgentiluomo/how-to-gwas-pdac/releases/download/v0.1-data/"
for (f in c("pdac_demo.bed", "pdac_demo.bim", "pdac_demo.fam")) {
  download.file(paste0(base, f), destfile = f, mode = "wb")
}
list.files(pattern = "pdac_demo")
```

### Command line

```bash
# macOS / Linux
for f in pdac_demo.bed pdac_demo.bim pdac_demo.fam; do
  curl -L -O "https://github.com/mgentiluomo/how-to-gwas-pdac/releases/download/v0.1-data/$f"
done
```

```powershell
# Windows PowerShell
$base = "https://github.com/mgentiluomo/how-to-gwas-pdac/releases/download/v0.1-data/"
foreach ($f in "pdac_demo.bed","pdac_demo.bim","pdac_demo.fam") {
  Invoke-WebRequest -Uri "$base$f" -OutFile $f
}
```

:::

::: {.callout-tip title="Pick one working folder and stay in it"}
Create a single folder for this tutorial — for example `gwas_tutorial` — and put the three data
files there. Do all your work inside it. This avoids almost every "file not found" problem
beginners run into.
:::

### Check the files arrived

In your terminal, move into the folder and list its contents:

```bash
cd path/to/gwas_tutorial      # replace with the real path to your folder
ls pdac_demo.*                # on Windows Command Prompt use: dir pdac_demo.*
```

You should see exactly three files: `.bed`, `.bim` and `.fam`.

---

## Step 2 — Install PLINK 2 {#sec-plink}

**PLINK** is the workhorse program for genotype data. We use **PLINK 2** (the command is
`plink2`). It is a single program file — there is **no installer**: you download it, unzip it,
and you are done.

1. Go to the official download page: <https://www.cog-genomics.org/plink/2.0/>
2. Download the build for your operating system (Linux / macOS / Windows).
3. **Unzip** it. Inside you will find the program file: `plink2` (macOS/Linux) or
   `plink2.exe` (Windows). Put it in your `gwas_tutorial` folder, next to the data.

::: {.callout-important title="How to call plink2 on your system"}
How you type the command depends on where PLINK lives and which system you use:

- **Windows (PowerShell):** write **`.\plink2`** — the `.\` in front is **required**, otherwise
  you get *"plink2 is not recognized…"*.
- **macOS / Linux**, with PLINK in your working folder: write **`./plink2`**.
- **Any system**, if PLINK is installed system-wide (on your `PATH`): just **`plink2`**.

For brevity, the rest of the guide writes the command simply as `plink2`. **Add the right prefix
for your system** (`.\` on Windows, `./` on macOS/Linux) whenever PLINK sits in your working
folder. Everything *after* the program name is identical on every system.
:::

::: {.panel-tabset}

#### Windows

Open *PowerShell*, move to your folder, and call PLINK with `.\` in front:

```powershell
cd path\to\gwas_tutorial
.\plink2 --version
```

#### macOS / Linux

Call it with `./` in front. The first time on macOS you may also need to allow it to run:

```bash
chmod +x plink2          # make the file executable (macOS/Linux only)
./plink2 --version
```

:::

---

## Step 3 — Check that PLINK runs {#sec-version}

Before doing anything with the data, confirm PLINK starts at all:

```bash
plink2 --version          # Windows: .\plink2 --version    macOS/Linux (local file): ./plink2 --version
```

You should see a single line with the version and build date, similar to:

```
PLINK v2.0.0-a.7.1 64-bit (4 May 2026)
```

If you see a version line, PLINK works. If instead you get an error like
`command not found` or `'plink2' is not recognized`, jump to
[Troubleshooting](#sec-troubleshooting) below.

---

## Step 4 — Your first command: allele frequencies with `--freq` {#sec-freq}

Now the real test. We ask PLINK to read the demo dataset and compute the **allele frequency** of
every variant — that is, how common each allele is across the sample. This is the simplest
possible operation that touches the actual genotypes, so it is a perfect "does everything work?"
check.

```bash
plink2 --bfile pdac_demo --freq --out pdac_demo_freq
```

*(Remember the prefix for your system: `.\plink2 …` on Windows, `./plink2 …` on macOS/Linux —
see Step 2.)*

Let's read that command piece by piece:

| Part | Meaning |
|------|---------|
| `plink2` | run the PLINK 2 program |
| `--bfile pdac_demo` | load the dataset whose three files share this name (`.bed`/`.bim`/`.fam`) |
| `--freq` | compute the allele frequencies |
| `--out pdac_demo_freq` | name the output files with this prefix |

<span class="codetag">CODE 0-01: first PLINK run — allele frequencies — see /getting_started/</span>

When it finishes, PLINK prints a short log to the screen and creates a new file:
**`pdac_demo_freq.afreq`**. The on-screen log ends with something like:

```
1461 samples (712 females, 749 males; 1461 founders) loaded from pdac_demo.fam.
430000 variants loaded from pdac_demo.bim.
Note: No phenotype data present.
Calculating allele frequencies... done.
--freq: Allele frequencies (founders only) written to pdac_demo_freq.afreq .
```

If you see the line saying the allele frequencies were **written**, the test has passed: PLINK can
read your data and compute on it. 🎉

(The "founders only" note simply means PLINK used the unrelated individuals for the frequency
calculation — exactly what we want.)

---

## Step 5 — Did it work? Look at the result {#sec-check}

Open the first few lines of the output file to see what PLINK produced:

```bash
head pdac_demo_freq.afreq            # Windows PowerShell: Get-Content pdac_demo_freq.afreq -Head 10
```

You should see a table with one row per variant, like this:

```
#CHROM  ID              REF  ALT  PROVISIONAL_REF?  ALT_FREQS   OBS_CT
1       1:817341:A:G    A    G    Y                 0.303854    2906
1       1:858952:G:A    G    A    Y                 0.119573    2902
1       1:899450:G:A    G    A    Y                 0.0107044   2896
1       1:903175:C:A    C    A    Y                 0.298898    2904
...
```

What the columns mean:

| Column | Meaning |
|--------|---------|
| `#CHROM` | chromosome |
| `ID` | the variant's identifier |
| `REF` / `ALT` | the reference and alternate alleles |
| `PROVISIONAL_REF?` | `Y` means the REF allele was assigned by PLINK rather than from a reference genome — normal here, and it does not affect the frequencies |
| `ALT_FREQS` | frequency of the **ALT** allele (a number between 0 and 1) |
| `OBS_CT` | number of observed alleles (≈ 2 × number of genotyped individuals) |

That `ALT_FREQS` column is the answer to our test: **PLINK computed a frequency for every
variant.**

### Optional: read it in R

If you have R/RStudio set up, you can load the same file and take a quick look. This also
introduces the pattern we use throughout the guide — PLINK produces a table, R reads and
summarises it:

```{r}
#| eval: false
# read the allele-frequency table PLINK just wrote
freq <- read.table("pdac_demo_freq.afreq",
                   header = TRUE, comment.char = "", sep = "\t",
                   check.names = FALSE)

# how many variants, and the first rows
nrow(freq)
head(freq)

# minor allele frequency (MAF) = the smaller of ALT_FREQS and 1 - ALT_FREQS
freq$MAF <- pmin(freq$ALT_FREQS, 1 - freq$ALT_FREQS)

# a quick picture of the frequency spectrum
hist(freq$MAF, breaks = 50,
     main = "Minor allele frequency distribution",
     xlab = "MAF")
```

A typical genotyping array shows many common variants and a tail of rarer ones, so the histogram
leans toward the lower MAF values. If R prints a number of variants and draws a histogram, your
R setup works too — and you are ready for the analysis sections.

::: {.callout-note title="Why we don't re-run PLINK every time the website is built"}
Real GWAS commands (especially on full datasets) can be slow and run on a dedicated server.
In this guide, PLINK commands are shown for **you** to run on your own machine; the website does
not re-execute them. Lightweight R steps (a plot, a small table) are run once and cached, so the
site stays fast and reproducible.
:::

---

## Troubleshooting {#sec-troubleshooting}

::: {.callout-warning collapse="true" title="`'plink2' is not recognized` / `command not found`"}
The terminal cannot find the PLINK program — usually because you forgot the prefix, or you are
not in the folder where PLINK lives.

- **Windows PowerShell:** call it as **`.\plink2`** (note the `.\`). This is the most common cause.
- **macOS / Linux:** call it as **`./plink2`** (note the `./`).
- Make sure you have `cd`-ed into your working folder.
- Check the file is really there: `dir plink2.exe` (Windows) or `ls plink2` (macOS/Linux).
:::

::: {.callout-warning collapse="true" title="`Error: Failed to open pdac_demo.bed`"}
PLINK cannot find the data files.

- Confirm all **three** files (`.bed`, `.bim`, `.fam`) are in the current folder and share the
  exact name `pdac_demo`.
- Confirm you typed the name **without** an extension after `--bfile` (PLINK adds `.bed`/`.bim`/`.fam`
  itself).
- Check the browser did not add a hidden `.txt` to a file name (e.g. `pdac_demo.fam.txt`).
:::

::: {.callout-warning collapse="true" title="macOS: \"plink2 cannot be opened because the developer cannot be verified\""}
This is macOS Gatekeeper. Either right-click the `plink2` file → *Open* once to allow it, or run
`xattr -d com.apple.quarantine plink2` in the terminal, then try again.
:::

---

## What's next

You now have a working setup and have confirmed PLINK can read the demo data. The next sections
of the guide build directly on this:

- **Quality control** — cleaning the raw genotypes before any analysis.
- **Population stratification** — detecting and correcting for ancestry.

*(These sections will be linked here as they are added.)*
